---
title: "Ch_16_Dates_and_Times"
author: "Samir Gadkari"
date: "1/29/2021"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)

read_path = Sys.getenv('DATASETS')
write_path = paste(read_path, 'output', sep = '/')
working_dir = Sys.getenv('R_WORKING_DIR')

full_read_path <- function(filename) {
  paste(read_path, filename, sep = '/')
}

full_write_path <- function(filename) {
  paste(write_path, filename, sep = '/')
}

full_wd_path <- function(filename) {
  paste(working_dir, filename, sep = '/')
}
```

Dates and times seem easy, but as you learn more they get more and
more complicated. Some of this is because:
  
  * many parts of the world use daylight savings time (DST), 
    so that some days have 23 hours, and others have 25.
  * some minutes have 61 seconds because every now and then leap seconds
    are added because the Earth’s rotation is gradually slowing down.

We will use the lubridate package to help with dates and datetimes.
R does not have a native class for storing times.
If you want to learn about times, look at the hms package.

## 16.2 Creating dates and times

Tibbles prints dates/times/datetimes differently:

  * dates: printed as <date>
  * times: printed as <time>
  * datetime: printed as <dttm> (actually, I see this printed as POSIXct).
    These are called POSIXct elsewhere in R.
  
Always use the simplest type possible. Datetimes are much more complicated
than dates, because of time zones.

Get the current date or datetime:
```{r}
today()
now()
```

You can create date/time from:

  * a string
  * individual date-time components
  * existing date/time object
  
### 16.2.1 From strings

To create dates from strings, give the ymd, mdy, dmy functions a string
with the correct order of components. These functions also take
unquoted numbers.
```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

ymd(20170131)
mdy(01312017)
dmy(31012017)
```

If the string cannot be parsed, an error will be returned.

To create a datetime, to the above function names, add an underscore and
one or more of "h", "m", "s":
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("Jan 31 2017 08:01")
```

Force the creation of a datetime from a date by supplying a timezone:
```{r}
ymd("20170131", tz = "UTC")
```

### 16.2.2 Individual components

  * make_datetime(year, month, day, hour, min) to create datetime from
    individual components
  * make_date(year, month, day) to create date from individual components
```{r}
flights %>%
  select(year, month, day, hour, minute)
```
  
Notice that year, month, day are integers and hour, minute are doubles.
```{r}
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

Let's do the same for our other time columns where the times are
in an odd format:
```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>%
  filter(!is.na(dep_time), !is.na(arr_time)) %>%
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>%
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```

Let's see the number of departures for the whole year.
Since datetimes have  a resolution of 1 second, binwidth is
given in seconds.

```{r}
flights_dt %>%
  ggplot(aes(dep_time)) +
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```
Or within a single day:
```{r}
flights_dt %>%
  filter(dep_time < ymd(20130102)) %>%
  ggplot(aes(dep_time)) +
  geom_freqpoly(binwidth = 600) # 600 seconds = 10 minutes
```

### 16.2.3 From other types

  * as_datetime: converts date to datetime.
    If the numeric offset from unix epoch (1970-01-01) is given
    in seconds, it can also convert it to datetime.
  * as_date:     converts datetime to date
    If the numeric offset from unix epoch (1970-01-01) is given
    in days, it can also convert it to date
  
```{r}
as_datetime(today())
as_date(now())
```
  
```{r}
as_datetime(60 * 60 * 60)
as_date(365 * 10 + 2)
```

## 16.2.4 Exercises

1. What happens if you parse a string that contains invalid dates?
```{r}
ymd(c("2010-10-10", "bananas"))
```
You get an error message "failed to parse" along with an NA value
for the element that could not be parsed.

2. What does the tzone argument to today() do? Why is it important?

The tzone argument is a character specifying which time zone you would
like the current time in. It defaults to your computer's system timezone.
It is important because it provides a frame of reference for time
comparisons. You can compare times in the same time zones easily.
You can compare times in different time zones if you know the offset
of the time zones.

3. Use the appropriate lubridate function to parse each of the following dates:
```{r}
d1 <- "January 1, 2010"
mdy(d1)

d2 <- "2015-Mar-07"
ymd(d2)

d3 <- "06-Jun-2017"
dmy(d3)

d4 <- c("August 19 (2015)", "July 1 (2015)")
mdy(d4)

d5 <- "12/30/14" # Dec 30, 2014
mdy(d5)
```

## 16.3 Date-time components

Now that we can get datetime into R's datetime data structures,
let's see how we can access parts of that structure

### 16.3.1 Getting components

All of these functions can be used to extract parts of a datetime
structure. They all return a number by default:

  * year
  * month: set label = TRUE to return the abbreviated name of the month.
    Set label = TRUE & abbr = FALSE to return the full name of the month.
  * mday:  get the day of the month from structure
  * yday:  get the day of the year from structure
  * wday:  get the day of the week from structure.
    Set label = TRUE to return the abbreviated name of the weekday.
    Set label = TRUE & abbr = FALSE to return the full name of the weekday
  * hour
  * minute
  * second
  
```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```
```{r}
month(datetime)
month(datetime, label = TRUE)
month(datetime, label = TRUE, abbr = FALSE)
```

```{r}
wday(datetime)
wday(datetime, label = TRUE)
wday(datetime, label = TRUE, abbr = FALSE)
```

We can use wday to see more flights depart during the week than on the
weekend:
```{r}
flights_dt %>%
  mutate(day = wday(dep_time, label = TRUE, abbr = FALSE)) %>%
  ggplot(aes(day)) +
  geom_bar()  # same as geom_histogram(stat = "count")
```

There's an interesting pattern if we look at average departure delay
for minute within the hour:
```{r}
flights_dt %>%
  mutate(minute = minute(dep_time)) %>%
  group_by(minute) %>%
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n()) %>%
  ggplot(aes(minute, avg_delay)) +
  geom_line()
```
The average delay seems to be less from 20 to 30 minutes,
and from 50 to 60 minutes.

If we look at the scheduled departure time, there is no such pattern:
```{r}
sched_dep <- flights_dt %>%
  mutate(minute = minute(sched_dep_time)) %>%
  group_by(minute) %>%
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n())

sched_dep %>%
  ggplot(aes(minute, avg_delay)) +
  geom_line()
```

So why is there such a pattern with actual departure times?
When humans are involved, there is a strong bias for flights leaving
at "nice" departure times. Always be aware of bias among humans.
```{r}
sched_dep %>%
  ggplot(aes(minute, n)) +
  geom_line()
```

### 16.3.2 Rounding

