---
title: "Ch_18_pipes"
author: "Samir Gadkari"
date: "2/2/2021"
output: html_document
---

Pipes (%>%, etc.) come with the magrittr package which is a part of
tidyverse. We will see how it works by exploring just the magrittr package,
instead of the entire tidyverse.

```{r}
library(magrittr)
```

## 18.2 Piping alternatives

Suppose you had to code the little bunny song:

    Little bunny Foo Foo
    Went hopping through the forest
    Scooping up the field mice
    And bopping them on the head

You could:

  1. Create a function for each step, saving the output and giving it
     to the next function.
     
    1.1 This requires a lot of names to be assigned to the intermediate
        variables.

    1.2 You may make a mistake and pass the wrong variable to the next
        function

  2. Create a function for each step, replacing the input with the output
     at each step, and passing the output to the next function
     
    2.1 If you had to debug your code, you would have to re-create the
        original variable again (since it has been replaced with 
        different values)
         
  3. Compose functions
  
    3.1 Since the inner function is used first, you have to read your code
        from the inside-out

  4. Use pipes
  
    4.1 Pipes take the computation on the left and pass it as input to
        the right. For the bunny exercise, magrittr creates something
        like this:

    my_pipe <- function(.) {
      . <- hop(., through = forest)
      . <- scoop(., up = field_mice)
      , <- bop(., on = head)
    }
    
    my_pipe(foo_foo)

        
    4.2 Pipes cannot be used with functions that use the 
        current environment.
```{r}
assign("x", 10)
x
"x" %>% assign(100)
x
```
This doesn't work because pipes assign the output of "x" %>% to a temporary
variable. Although this variable is changed to 100, the function pipe
has created to handle piping does not return the updated value out.
Instead, explicitly use an environment:
```{r}
env <- environment()
"x" %>% assign(100, envir = env)
x
```

Other functions with this problem: get() and load()

    4.3 The pipe computes each section in turn, passing the value forward.
        This is why functions that use lazy eval cannot be used with pipes.

One example is tryCatch:
```{r}
tryCatch(stop("!"), error = function(e) "An error")

# stop("!") %>%
#   tryCatch(error = function(e) "An error")
# Returns "Error in eval(lhs, parent, parent) : !
```

There are a relatively wide class of functions with this behavior, ex.
try, suppressMessages, suppressWarnings.

## 18.3 When not to use pipe

  * If your pipes are more than, say 10 steps long. Instead, use
    intermediate variables that increase understanding of the code.
  * If you have multiple inputs or outputs
  * Your code will be a directed graph with a complex dependency structure
  
## 18.4 Other tools from magrittr

  * Sometimes, you want to print out a value computed within the pipe (or
    plot it, or save it to disk). You can put your plot function within
    the pipe structure, prepending it with a %T>% (Tee). This tells 
    magrittr to plot the graph, but not propagate the output of plot. 
    Instead, propagate the output of the code before the %T>%:
```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%  # You will see this plot.
  str()       # Output of the str functions is NULL

# Instead, do:
rnorm(100) %>%
  matrix(ncol = 2) %T>% # Propagates output of matrix ...
  plot() %>%
  str()                 # ... to the input of str.
```
    
  * When you want to apply a function to variables, instead of the whole
    dataframe, use %$%. It "explodes" the dataframe into variables:
```{r}
mtcars %$%
  cor(disp, mpg)
```
    
  * For assignment, you can use %<>%.
Instead of
```{r}
mtcars <- mtcars %>%
  transform(cyl = cyl * 2)
```

do:
```{r}
mtcars %<>% transform(cyl = cyl * 2)
```

Still, the earlier approach is better because it is more explicit.