---
title: "Ch_10_Tibbles"
author: "Samir Gadkari"
date: "1/17/2021"
output: html_document
---
```{r}
library(tidyverse)
```

A tibble is like a data frame, but makes life easier by:
  * never changes an input's type
  * never changes the names of variables.
    ex. `crazy name` stays as 'crazy name', not 'crazy.name'
    like with a data.frame.
  * it never creates row names
  * it evaluates it's arguments lazily and sequentially.
    You can do: tibble(x = 1:5, y = x ^ 2)
  * it never uses row.names(). It stores data in a consistent
    way - never stores variable as a special attribute
  * only recycles vectors of length 1. Recycling vectors of
    greater lengths is a frequent source of bugs.

## Creating tibbles
You can use as_tibble(x) to convert data.frame x to a tibble.
as_tibble() is performant.
```{r}
as_tibble(iris)
```

Create a new tibble from individual vectors with tibble():
```{r}
tibble(x = 1:5,
       y = 1,  # tibble conversion only recycles vectors of len 1
       z = x ^ 2 + y)
```

A tibble variable can have names that R variables cannot.
Create them surrounded by backticks.
You will need to use backticks in other packages (ggplot2,
dplyr, tidyr) when referring to them.
```{r}
tb <- tibble(
  `:)` = 'smile',
  ` ` = 'space',
  `2000` = 'number'
)
```
You can create a tibble with data entry in code using tribble():
```{r}
(x <- tribble(
  ~x, ~y, ~z, # Column headings defined by formulas
              # Formulas show up as x, y, z
  'a', 2.0, 3.6,
  'b', 1, 8.5
))
```

## 10.3.1 Printing
Tibbles are printed as 10 rows with as many columns as can fit.
The output include the class of the variable.
```{r}
tibble(
  # runif(n) creates a normal distribution of n values in the
  # 0.0 - 1.0 range
  a = lubridate::now() + runif(1e3) * 86400, # 86400 sec = 24 hrs
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

To print more than 10 lines or more columns:
```{r}
nycflights13::flights %>% 
  print(n = 10,       # 10 rows
        width = Inf)  # All columns
```
You can use options so that tibbles are always printed in a
particular way:

options(tibble.print_max = n, # Print max n rows.
        tibble.print_min = m) # print min m rows if >m rows

options(tibble.print_min = Inf) # print all rows

options(tibble.width = Inf) # print all columns

All options are listed at package?tibble

To see the full tibble, use View(tb) in RStudio

## 10.3.2 Subsetting

```{r}
(df <- tibble(
  x = runif(5),
  y = rnorm(5)
))

df$x      # extract by name
df[['x']] # extract by name

df[[1]]   # extract by position

# To use in a pipe, use the special placeholder '.'
df %>% .$x
df %>% .[['x']]
df %>% .[[1]]
```

## 10.4 Interacting with older code

Some older functions don't work with tibbles.
Use as.data.frame() to convert to a data frame,
use the function, then as_tibble() to convert back to tibble.

With base R data frames, the [ function sometimes returns
a data frame and sometimes a vector.
With tibbles, the [ function always returns a tibble.

## 10.5 Exercises

1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).
```{r}
print(mtcars)
```
You can see that this data.frame is printed with all columns.
A tibble is printed with just enough columns to fill the screen.

2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?
```{r}
(df <- data.frame(abc = 1, xyz = "a"))
(df$x)
(df[, "xyz"])
(df[, c("abc", "xyz")])
```
The df$x (because x is the start of the second column name),
prints the column out. We would like it to print an error,
because there may be another column whose name started with x
that we're interested in printing. We may end up getting a
wrong result somewhere further down the code, and have to
debug it. This is unnecessary, if there is an error instead.

3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?
```{r}
var <- 'mpg'
# mtcars$var      # This does not work
mtcars[[var]]   # but this does.
```

4. Practice referring to non-syntactic names in the following data frame by:
    Extracting the variable called 1.
    Plotting a scatterplot of 1 vs 2.
    Creating a new column called 3 which is 2 divided by 1.
    Renaming the columns to one, two and three.
```{r}
(annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
))

# annoying[[`1`]]  # This does not work
annoying$`1`     # but this does
annoying[[1]]    # This works, but here we're referring 
                 # to it by location

annoying %>%
  ggplot(aes(`1`, `2`)) +
  geom_point()

annoying %>%
  mutate(`3` = `2` / `1`) %>%
  rename('one' = `1`,
         'two' = `2`,
         'three' = `3`)
```

5. What does tibble::enframe() do? When might you use it?
enframe converts vectors or lists into 2-column tibbles.
By default, the first column is named 'name', and the
second column is named 'value'. You can change this using
name and value arguments to enframe. enframe is useful in
constructing small tibbles to test with.

deframe converts 2-column data frames to a named vector or list.

6. What option controls how many additional column names are printed at the footer of a tibble?

tibble.max_extra_cols
