---
title: "Ch_16_Dates_and_Times"
author: "Samir Gadkari"
date: "1/29/2021"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)

read_path = Sys.getenv('DATASETS')
write_path = paste(read_path, 'output', sep = '/')
working_dir = Sys.getenv('R_WORKING_DIR')

full_read_path <- function(filename) {
  paste(read_path, filename, sep = '/')
}

full_write_path <- function(filename) {
  paste(write_path, filename, sep = '/')
}

full_wd_path <- function(filename) {
  paste(working_dir, filename, sep = '/')
}
```

Dates and times seem easy, but as you learn more they get more and
more complicated. Some of this is because:
  
  * many parts of the world use daylight savings time (DST), 
    so that some days have 23 hours, and others have 25.
  * some minutes have 61 seconds because every now and then leap seconds
    are added because the Earth’s rotation is gradually slowing down.

We will use the lubridate package to help with dates and datetimes.
R does not have a native class for storing times.
If you want to learn about times, look at the hms package.

## 16.2 Creating dates and times

Tibbles prints dates/times/datetimes differently:

  * dates: printed as <date>
  * times: printed as <time>
  * datetime: printed as <dttm>
    these are called posixct elsewhere in R.
  
Always use the simplest type possible. Datetimes are much more complicated
than dates, because of time zones.

Get the current date or datetime:
```{r}
today()
now()
```

You can create date/time from:

  * a string
  * individual date-time components
  * existing date/time object
  
### 16.2.1 From strings

To create dates from strings, give the ymd, mdy, dmy functions a string
with the correct order of components. These functions also take
unquoted numbers.
```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

ymd(20170131)
mdy(01312017)
dmy(31012017)
```

If the string cannot be parsed, an error will be returned.

To create a datetime, to the above function names, add an underscore and
one or more of "h", "m", "s":
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("Jan 31 2017 08:01")
```

Force the creation of a datetime from a date by supplying a timezone:
```{r}
ymd("20170131", tz = "UTC")
```

