---
title: "Ch_15_Factors"
author: "Samir Gadkari"
date: "1/27/2021"
output: html_document
---

```{r}
library(tidyverse)

read_path = Sys.getenv('DATASETS')
write_path = paste(read_path, 'output', sep = '/')
working_dir = Sys.getenv('R_WORKING_DIR')

full_read_path <- function(filename) {
  paste(read_path, filename, sep = '/')
}

full_write_path <- function(filename) {
  paste(write_path, filename, sep = '/')
}

full_wd_path <- function(filename) {
  paste(working_dir, filename, sep = '/')
}
```

We will use the forcats (anagram of factors) package in tidyverse
to work with factors.

Factors are useful when:

  * working with categorical variables
  * when you have to display character vectors in a non-alphabetical order
  
## 15.2 Creating factors

If you used character vectors to list the months of the year,
you must spell each month the same way each time. If you have spelling
mistakes, you will get different results.
Also, you can't sort characters vectors Jan through Dec !

Let's create a factor:
```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```
Any values not in the set will be silently converted to NA:
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
y2 <- factor(x2, levels = month_levels)
y2
```
To get a warning, use readr::parse_factor. This is really useful when
testing your variables for unusual data.
```{r}
y2 <- parse_factor(x2, levels = month_levels)
```

If you omit the levels, they will be taken in alphabetical order from the
character vector:
```{r}
factor(x1)
```

To set levels according to first-seen levels in the data, you can use
either:

  * unique(data)
  * or get default factoring, and then use fct_inorder(data)
```{r}
f1 <- factor(x1, levels = unique(x1))
f1
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```
levels function gives you the levels of a factor:
```{r}
levels(f2)
```

## General Social Survey

For the rest of this chapter we will use the forcats:gss_cat dataset.
It is a social survey with a lot of variables. We have only selected
a few of them. It was created by [NORC]('http://gss.norc.org/').
The survey was done over the years 2000-2014.

```{r}
gss_cat
```

To see the levels of a factor, use count:
```{r}
gss_cat %>% count(race)
```
or with a bar chart.
By default, geom_bar will drop counts that are 0.
Use scale_x_discrete(drop = FALSE) to show all factors, even those with
0 counts.
Unfortunately, dplyr doesn't have a drop option, but will have
in the future.
```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar()

ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

### 15.3.1 Exercise

1. Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

The default bar chart is difficult to read because the values on the x
axis overlap. Use coord_flip to put them on the y axis.
```{r}
ggplot(gss_cat, aes(rincome)) +
  geom_bar() +
  coord_flip() # %>%
  # scale_x_discrete(drop = FALSE)
```

2. What is the most common relig in this survey? What’s the most common partyid?

```{r}
gss_cat %>%
  group_by(relig) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(1) %>%
  pull(relig)

gss_cat %>%
  group_by(partyid) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(1) %>%
  pull(partyid)
```

3. Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r}
gss_cat %>%
  select(relig, denom) %>%
  group_by(relig, denom) %>%
  summarise(num_denom = n()) %>%
  pivot_wider(names_from = relig, values_from = num_denom)

gss_cat %>%
  select(relig, denom) %>%
  group_by(relig, denom) %>%
  summarise(num_denom = n()) %>%
  ggplot(aes(relig, denom)) +
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1.1, 
                                   hjust = 1.2))
  
```
As you look at the numbers, you can see that most of the denominations
are for the Protestants.

## 15.4 Modifying factor order

It's often useful to change the order of the factor levels.
Let's say you want to explore the average number of hours spent watching
TV per day per religion
```{r}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()
```
There is no pattern here. Reorder using fct_reorder. It takes arguments:

  * f, the factor you want to modify
  * x, the numeric vector you want to use for the modification
  * fun, optional function to use if there are multiple values of x
    for the same f. Default = median
    
```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```
Now you can easily see which religions watch how much TV.

With complicated transformations, it's better to move them to a
mutate step before using ggplot.
```{r}
relig_summary %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
  geom_point()
```
How about a similar plot with average age and income levels?
```{r}
rincome_summary <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) +
  geom_point()
```
Look at our reordered income levels. They're arbitrarily ordered.
So in this case, we should not have reordered them. Only use fct_reorder
for factors that are arbitrarily ordered.

Some of the levels like "Not applicable", "Refused", "Don't know", and
"No answer" can be brought to the bottom of the plot. You can use
fct_relevel for that. It 

  * fct_relevel: allows you to move any number of levels to any location.
    Takes a factor variable, the levels to move, and the location after
    which to put them. The order in which you specify the levels is how
    it shows up in the graph. If you don't specify a location to put
    them, they're put at the beginning (bottom of graph).

```{r}
# Without any adjustment to levels, the default graph looks like:
ggplot(rincome_summary, aes(age, rincome)) +
  geom_point()

# With the "Not applicable" level relevelled -
# by default, it's moved to the bottom of the plot.
# The location of the levels not mentioned are not changed.
ggplot(rincome_summary, 
       aes(age, fct_relevel(rincome, 
                            c("Not applicable")))) +
  geom_point() +
  ylab("rincome with relevelling")
```

The average age for "Not applicable" is high because those people are
retired.

Another type of reordering is useful when you are coloring the lines
on a plot. fct_reorder2 reorders the factor by the y values associated
with the largest x values.

  * fct_reorder2: takes a factor (.f), and two numerical 
    variables (.x, .y), and reorders the factor so that the values of the
    function .fun(.x, .y) are in ascending order. If there are multiple
    y values for the same x values, all the factors associated with
    that largest x value comes at the top of the legend in the plot.
    Then the set of factors associated with the second highest x value
    comes next, etc.
    You can pass na.rm = TRUE to the function. By default, .fun = last2. 
    
    + last2 sorts by .x and finds the last value of .y after the 
      sorting is done.
```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%  # Just like group_by, we get a number for each
                           # separate combination of age and marital
                           # values. The resulting variable is named n
  group_by(age) %>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, color = marital)) +
  geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(na.rm = TRUE)
```

  * fct_infreq: orders levels in increasing frequency (for bar plots).
    Orders the factor with the highest number of observations at the
    left of the bar chart.
  * fct_rev: reverses the order of the factor
```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq(marital) %>% fct_rev()) %>%
  ggplot(aes(marital)) +
  geom_bar()
```

### 15.4.1 Exercises

1. There are some suspiciously high numbers in tvhours. Is the mean a good summary?
```{r}
gss_cat %>%
  ggplot(aes(tvhours)) +
  geom_histogram(na.rm = TRUE)

median(gss_cat$tvhours, na.rm = TRUE)
```

The tvhours histogram has a long tail with a maximum value of 15.
This will skew the mean to the right. The median (value of 2) would be a better measure of tvhours watched by the sample.

2. For each factor in gss_cat identify whether the order of the levels is arbitrary or principled.

```{r}
# The factors in css_cat are:
# marital, race, rincome, partyid, relig, denom
# Also, for levels, the highest level is first, and then lesser
# and lesser levels.
levels(gss_cat$marital) # Never married should be after Married.
                        # Don't know where "No answer" should go.
                        # I was going to put it to the right, 
                        # since it's not useful, but just not answering
                        # a question can tell us something, maybe?
levels(gss_cat$race) # Race should not be ordered anyway, so this order
                     # is fine
levels(gss_cat$rincome) # Think this is fine ordering
levels(gss_cat$partyid) # Think this is fine ordering
levels(gss_cat$relig) # I think these levels should change.
                      # Christian, and Orthodox-christian are separated
                      # from Catholic and Protestant.
levels(gss_cat$denom) # I think the Presbyterians, Lutherans, Methodists,
                      # and Baptists are grouped near each other. This
                      # is good. What about how these groups should
                      # be in relation to each other? I don't know.
```

3. Why did moving “Not applicable” to the front of the levels move it to the bottom of the plot?

By default fct_relevel moves the selected levels to the first position -
which is at the bottom of the plot

## 15.5 Modifying factor levels

  * fct_recode: will change the values (names) of the levels of a factor.
    This allows you to group similar levels together, and clarify the
    levels for publication
    
```{r}
gss_cat %>% count(partyid)
```

Let's recode these levels:
```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",  # Only the levels
      "Republican, weak"      = "Not str republican", # we want to recode
      "Independent, near rep" = "Ind,near rep",       # are on the RHS.
      "Independent, near dem" = "Ind,near dem",       # LHS are the new
      "Democrat, weak"        = "Not str democrat",   # names for the same
      "Democrat, strong"      = "Strong democrat"     # levels.
    )
  ) %>%
  count(partyid)
```

If a level is named on the RHS that is not present, 
you will get a warning.
Combine levels by assigning multiple old levels to the same
new/existing level:
```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",
      "Republican, weak"      = "Not str republican",
      "Independent, near rep" = "Ind,near rep",
      "Independent, near dem" = "Ind,near dem",
      "Democrat, weak"        = "Not str democrat",
      "Democrat, strong"      = "Strong democrat",
      "Other"                 = "No answer",
      "Other"                 = "Don't know",
      "Other"                 = "Other party"
    )
  ) %>%
  count(partyid)
```

To combine lots of levels together, use fct_collapse. You can then define
multiple levels on a single line:
```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(
    partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

To lump together all the small groups to make the table simpler,
use fct_lump. fct_lump is now only used for historical reasons.
The recommendation is that you instead use one of:

  * fct_lump_min: lumps levels that appear < min times
  * fct_lump_prop: a positive prop lumps levels that appear with < prop.
    a negative prop lumps levels that do not appear with < -prop
  * fct_lump_n: lumps all levels except for n most frequent (or
    least frequent if n < 0). If there are ties, you will get at least
    abs(n) values.
  * fct_lump_lowfreq: lumps least frequent levels, ensuring that
    other_level is still the smallest level

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
```
Looks like we have overcollapsed.

We can use n to specify the minimum number of groups we will get:
```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10)) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

### 15.5.1 Exercises

1. How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    republican = c("Strong republican", "Not str republican"),
    independent = c("Ind,near rep", "Independent", "Ind,near dem"),
    democrat = c("Not str democrat", "Strong democrat"))) %>%
  count(year, partyid) %>%
  group_by(year) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(year, prop, color = fct_reorder2(partyid, year, prop))) +
  geom_line(na.rm = TRUE)
```

2. How could you collapse rincome into a small set of categories?

You can look at the proportions of rincome in the data:
```{r}
gss_cat %>%
  count(rincome) %>%
  summarise(rincome = rincome,
            prop = n / sum(n))
```
Now you can specify a threshold to lump factor levels below it:
```{r}
levels(gss_cat$rincome)  # original levels
print("")
levels(fct_lump_prop(gss_cat$rincome, 0.02))  # levels after 0.02
                                              # proportion is tweaked.

```

