---
title: "Ch_15_Factors"
author: "Samir Gadkari"
date: "1/27/2021"
output: html_document
---

```{r}
library(tidyverse)

read_path = Sys.getenv('DATASETS')
write_path = paste(read_path, 'output', sep = '/')
working_dir = Sys.getenv('R_WORKING_DIR')

full_read_path <- function(filename) {
  paste(read_path, filename, sep = '/')
}

full_write_path <- function(filename) {
  paste(write_path, filename, sep = '/')
}

full_wd_path <- function(filename) {
  paste(working_dir, filename, sep = '/')
}
```

We will use the forcats (anagram of factors) package in tidyverse
to work with factors.

Factors are useful when:

  * working with categorical variables
  * when you have to display character vectors in a non-alphabetical order
  
## 15.2 Creating factors

If you used character vectors to list the months of the year,
you must spell each month the same way each time. If you have spelling
mistakes, you will get different results.
Also, you can't sort characters vectors Jan through Dec !

Let's create a factor:
```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```
Any values not in the set will be silently converted to NA:
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
y2 <- factor(x2, levels = month_levels)
y2
```
To get a warning, use readr::parse_factor. This is really useful when
testing your variables for unusual data.
```{r}
y2 <- parse_factor(x2, levels = month_levels)
```

If you omit the levels, they will be taken in alphabetical order from the
character vector:
```{r}
factor(x1)
```

To set levels according to first-seen levels in the data, you can use
either:

  * unique(data)
  * or get default factoring, and then use fct_inorder(data)
```{r}
f1 <- factor(x1, levels = unique(x1))
f1
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```
levels function gives you the levels of a factor:
```{r}
levels(f2)
```

## General Social Survey

For the rest of this chapter we will use the forcats:gss_cat dataset.
It is a social survey with a lot of variables. We have only selected
a few of them. It was created by [NORC]('http://gss.norc.org/').
The survey was done over the years 2000-2014.

```{r}
gss_cat
```

To see the levels of a factor, use count:
```{r}
gss_cat %>% count(race)
```
or with a bar chart.
By default, geom_bar will drop counts that are 0.
Use scale_x_discrete(drop = FALSE) to show all factors, even those with
0 counts.
Unfortunately, dplyr doesn't have a drop option, but will have
in the future.
```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar()

ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

### 15.3.1 Exercise

1. Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

The default bar chart is difficult to read because the values on the x
axis overlap. Use coord_flip to put them on the y axis.
```{r}
ggplot(gss_cat, aes(rincome)) +
  geom_bar() +
  coord_flip() # %>%
  # scale_x_discrete(drop = FALSE)
```

2. What is the most common relig in this survey? What’s the most common partyid?

```{r}
gss_cat %>%
  group_by(relig) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(1) %>%
  pull(relig)

gss_cat %>%
  group_by(partyid) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(1) %>%
  pull(partyid)
```

3. Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r}
gss_cat %>%
  select(relig, denom) %>%
  group_by(relig, denom) %>%
  summarise(num_denom = n()) %>%
  pivot_wider(names_from = relig, values_from = num_denom)

gss_cat %>%
  select(relig, denom) %>%
  group_by(relig, denom) %>%
  summarise(num_denom = n()) %>%
  ggplot(aes(relig, denom)) +
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1.1, 
                                   hjust = 1.2))
  
```
As you look at the numbers, you can see that most of the denominations
are for the Protestants.

