---
title: "Ch_25_Many_models"
author: "Samir Gadkari"
date: "2/16/2021"
output: html_document
---

```{r}
library(modelr)
library(tidyverse)
library(gapminder)
```

We will use 3 powerful ideas to work with many models:

  * Use simple models to understand complex datasets
  * Use list-columns to store arbitrary data structures in a dataframe.
    ex. store linear model in a column
  * Use the broom package to turn models into tidy data.
    Once converted, you can use all the techniques presented here with
    the model's tidy data
    
## 25.2 gapminder

The gapminder data summarises the progression of countries over time, looking at statistics like life expectancy and GDP.
```{r}
gapminder
```

We're only going to ask one question in this chapter:
How does life expectancy (lifeExp) change over time (year) for each 
country (country)?

```{r}
gapminder %>%
  ggplot(aes(year, lifeExp, group = country)) +
  geom_line(alpha = 1 / 3)
```

Many countries show increasing life expectancy. How do we find the ones that
are going down? Let's fit regression lines and look at residuals to pick
those out. For a single country:

```{r}
nz <- filter(gapminder, country == "New Zealand")
nz %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>%
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) +
  geom_line() +
  ggtitle("Linear trend + ")

nz %>%
  add_residuals(nz_mod) %>%
  ggplot(aes(year, resid)) +
  geom_line() +
  ggtitle("Remaining pattern")
```

How do we easily fit that model to every country?

### 25.2.1 Nested data

We can extract out the common code and use purrr::map.
Instead of repeating an action for each variable, we need to repeat the action
for each country (subset of rows). To do that we need to create a
nested data frame. Start with a grouped data frame, then nest it.

```{r}
by_country <- gapminder %>%
  group_by(country, continent) %>% # we want to carry continent along,
                                   # so we add it to groupby
  nest()
by_country
```

The last column that was created in that dataframe was a list of tibbles.
These are the embedded dataframes inside the dataframe.
You should not use str on the parent dataframe, because it creates
a long list of output.

The columns within each tibble are the columns associated
with that country.

```{r}
by_country$data[[1]]
```

We now have a meta-observation in the tibble: the complete
time course for that country.
In a grouped dataframe, each row is an observation.
In a nested dataframe, each row is a group.

### 25.2.2 List-columns

Let's make a model-fitting function, and apply it to each dataframe.
```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

models <- map(by_country$data, country_model)
```

Instead of keeping this list of models as a separate object,
we can keep it as a column in the by_country dataframe.
```{r}
by_country <- by_country %>%
  mutate(model = map(data, country_model))
by_country
```

This keeps all models together with the data, even when you
filter or arrange the parent dataframe.

```{r}
by_country %>%
  arrange(continent, country)
```

### 25.2.3 Unnesting

Let's compute the residuals for all the models:
```{r}
by_country <- by_country %>%
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```

```{r}
by_country$resids[[1]]
```

Now, you can't plot a list of dataframes. So, let's convert these back
to a single dataframe with unnest:

```{r}
resids <- unnest(by_country, resids)
resids
```

The resid tibbles in the dataframe are exploded and those columns 
show up as columns in the output. Each column is preceded by
repeated country and continent (all the columns that were used
to group the original dataframe - just as expected, since this
is the opposite of nest).

Now we can plot the residuals:

```{r}
resids %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1 / 3) +
  geom_smooth(se = FALSE)
```

Faceting by continents is particularly revealing:

```{r}
resids %>%
  ggplot(aes(year, resid, group = country)) +
  geom_line(alpha = 1 / 3) +
  facet_wrap(~ continent)
```

Looks like we're missing some mild patterns overall, 
that there are large residuals in Africa. We should
look there too.

### 25.2.4 Model quality

