---
title: "Ch_25_Many_models"
author: "Samir Gadkari"
date: "2/16/2021"
output: html_document
---

```{r}
library(modelr)
library(tidyverse)
library(gapminder)
```

We will use 3 powerful ideas to work with many models:

  * Use simple models to understand complex datasets
  * Use list-columns to store arbitrary data structures in a dataframe.
    ex. store linear model in a column
  * Use the broom package to turn models into tidy data.
    Once converted, you can use all the techniques presented here with
    the model's tidy data
    
## 25.2 gapminder

The gapminder data summarises the progression of countries over time, looking at statistics like life expectancy and GDP.
```{r}
gapminder
```

We're only going to ask one question in this chapter:
How does life expectancy (lifeExp) change over time (year) for each 
country (country)?

```{r}
gapminder %>%
  ggplot(aes(year, lifeExp, group = country)) +
  geom_line(alpha = 1 / 3)
```

Many countries show increasing life expectancy. How do we find the ones that
are going down? Let's fit regression lines and look at residuals to pick
those out. For a single country:

```{r}
nz <- filter(gapminder, country == "New Zealand")
nz %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>%
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) +
  geom_line() +
  ggtitle("Linear trend + ")

nz %>%
  add_residuals(nz_mod) %>%
  ggplot(aes(year, resid)) +
  geom_line() +
  ggtitle("Remaining pattern")
```

How do we easily fit that model to every country?

### 25.2.1 Nested data

We can extract out the common code and use purrr::map.
Instead of repeating an action for each variable, we need to repeat the action
for each country (subset of rows). To do that we need to create a
nested data frame. Start with a grouped data frame, then nest it.

```{r}
by_country <- gapminder %>%
  group_by(country, continent) %>% # we want to carry continent along,
                                   # so we add it to groupby
  nest()
by_country
```

The last column that was created in that dataframe was a list of tibbles.
These are the embedded dataframes inside the dataframe.
You should not use str on the parent dataframe, because it creates
a long list of output.

The columns within each tibble are the columns associated
with that country.

```{r}
by_country$data[[1]]
```

We now have a meta-observation in the tibble: the complete
time course for that country.
In a grouped dataframe, each row is an observation.
In a nested dataframe, each row is a group.

### 25.2.2 List-columns

Let's make a model-fitting function, and apply it to each dataframe.
```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

models <- map(by_country$data, country_model)
```

Instead of keeping this list of models as a separate object,
we can keep it as a column in the by_country dataframe.
```{r}
by_country <- by_country %>%
  mutate(model = map(data, country_model))
by_country
```

This keeps all models together with the data, even when you
filter or arrange the parent dataframe.

```{r}
by_country %>%
  arrange(continent, country)
```

### 25.2.3 Unnesting

Let's compute the residuals for all the models:
```{r}
by_country <- by_country %>%
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```

```{r}
by_country$resids[[1]]
```

Now, you can't plot a list of dataframes. So, let's convert these back
to a single dataframe with unnest:

```{r}
resids <- unnest(by_country, resids)
resids
```

The resid tibbles in the dataframe are exploded and those columns 
show up as columns in the output. Each column is preceded by
repeated country and continent (all the columns that were used
to group the original dataframe - just as expected, since this
is the opposite of nest).

```{r}
resids$data[[1]]
```

The data still contains the embedded tibble for each country.

Now we can plot the residuals:

```{r}
resids %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1 / 3) +
  geom_smooth(se = FALSE)
```

Faceting by continents is particularly revealing:

```{r}
resids %>%
  ggplot(aes(year, resid, group = country)) +
  geom_line(alpha = 1 / 3) +
  facet_wrap(~ continent)
```

Looks like we're missing some mild patterns overall, and
that there are large residuals in Africa. We should
look there too.

### 25.2.4 Model quality

The broom package provides some functions to turn models
into tidy data. broom::glance extracts model quality metrics:
```{r}
broom::glance(nz_mod)
```

```{r}
glance <- by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance)
glance
```

Now we start looking for models that don't fit well:

```{r}
glance %>%
  arrange(r.squared)
```

The worst-fit models appear to be all in Africa. Let's double check
with a plot. We will jitter the observations so they can all be seen.

```{r}
glance %>%
  ggplot(aes(continent, r.squared)) +
  geom_jitter(width = 0.5)
```

Let's look at the bad fit countries:

```{r}
bad_fit <- glance %>% filter(r.squared < 0.25)

gapminder %>%
  semi_join(bad_fit, by = "country") %>%
  ggplot(aes(year, lifeExp, color = country)) +
  geom_line()
```

These are the results of two tragedies: AIDS, and Rwandan genocide.

### 25.2.5 Exercises

1. A linear trend seems to be slightly too simple for the overall trend. Can you do better with a quadratic polynomial? How can you interpret the coefficients of the quadratic? (Hint you might want to transform year so that it has mean zero.)

The way to get a mean zero for the year is to subtract the mean
from each year value.

```{r}
country_model <- function(df) {
  lm(lifeExp ~ poly(mean_0_year + mean_0_year, 2), data = df)
}

by_country <- gapminder %>%
  group_by(country, continent) %>%
  mutate(mean_0_year = year - mean(year)) %>%
  nest()

by_country <- by_country %>%
  mutate(
    model = map(data, country_model),
    resids = map2(data, model, add_residuals)
  )

unnest(by_country, resids) %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1 / 3) +
  geom_smooth(se = FALSE)
```


Let's look at the residuals for Africa:
```{r}
glance <- by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance)

glance %>%
  arrange(r.squared) %>%
  filter(r.squared < 0.25)
```

Now there is only one country with R^2 < 0.25. It is Rwanda - which
must be because of the genocide.

Let's plot R^2 for each continent:

```{r}
glance %>%
  ggplot(aes(continent, r.squared)) +
  geom_jitter(width = 0.5)
```

2. Explore other methods for visualising the distribution of R2 per continent. You might want to try the ggbeeswarm package, which provides similar methods for avoiding overlaps as jitter, but uses deterministic methods.

```{r}
library(ggbeeswarm)

glance %>%
  ggplot(aes(continent, r.squared)) +
  geom_quasirandom(alpha = 1 / 3)
```

3. To create the last plot (showing the data for the countries with the worst model fits), we needed two steps: we created a data frame with one row per country and then semi-joined it to the original dataset. It’s possible to avoid this join. How?

All we needed from the join was lifeExp. Since we already have it in data,
we can unnest data and use it.
```{r}
glance %>%
  unnest(data) %>%
  filter(r.squared < 0.25) %>%
  ggplot(aes(year, lifeExp, color = country)) +
  geom_line()
```

## 25.3 List-columns

A data frame is a named list of equal length vectors.
A list is a vector, so it can be used as a column of a dataframe.
Base R does not allow you to create list-columns easily, and
data.frame treats a list as a list of columns:
```{r}
data.frame(x = list(1:3, 3:5))
```

You can prevent data.frame from doing this with I, but the
result doesn't print well:
```{r}
data.frame(x = I(list(1:3, 3:5)),
           y = c("1, 2", "3, 4, 5"))
```

Tibble alleviates this by being lazy, not modifying it's inputs,
and printing better:

```{r}
tibble(x = list(1:3, 3:5),
       y = c("1, 2", "3, 4, 5"))
```

Tribble is even easier, since it can automatically work out
that you need a list:

```{r}
tribble(
  ~x,    ~y,
  1:3,   "1, 2",
  3:5,   "3, 4, 5"
)
```

List-columns are most useful as an intermediate data structure.
Since base R works with atomic vectors and data frames,
it may be a hassle, but still worth it to keep data in
list-columns.

3 parts of an effective list-column pipeline:

  1. Create a list-column using:
     * nest
     * summarise + list
     * mutate + a map function
  2. Create other intermediate list-columns by transforming
     list-columns using map, map2, pmap
  3. Simplify the list-column back down to a dataframe
     or atomic vector
     
## 25.4 Creating list-columns

Typically you create list-columns using:

  1. tidyr::nest to convert a grouped dataframe into a nested
     dataframe where you have list-columns
  2. mutate and vectorised functions that return a list
  3. summarise and summary functions that return multiple results
You could also create them using tibble::enframe and a named list,
but this is not usual.

Generally, list-columns should have the same type for all items.

### 25.4.1 With nesting

nest() creates a nested dataframe where:

  * each row is a meta-observation
  * some columns give variables that define the observation
  * the list-column of dataframes gives the individual
    observations that make up the meta-observation

When nest is applied to grouped dataframes, it keeps the
grouping columns as is, and bundles everything into the list-column:

```{r}
gapminder %>%
  group_by(country, continent) %>%
  nest()
```

You can also apply nest to an ungrouped dataframe by specifying
which columns to nest:

```{r}
gapminder %>%
  nest(data = c(year : gdpPercap))
```

### 25.4.2 From vectorised functions

Some useful functions take an atomic vector and create a list.
ex. stringr::str_split. If this is used inside mutate,
you will get a list-column.

```{r}
df <- tribble(
  ~x1,
  "a, b, c",
  "d, e, f, g"
)

df %>%
  mutate(x2 = stringr::str_split(x1, ","))
```

unnest can help here:

```{r}
df %>%
  mutate(x2 = stringr::str_split(x1, ",")) %>%
  unnest(x2)
```

Check out tidyr::separate_rows which is a wrapper around this pattern.
It takes the data and a separator to separate the delimited values.
If convert = TRUE, it will even convert numeric/integer/logical values
to the appropriate type.

Another example of this is when using map, map2, and pmap from purrr:

```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)

sim %>%
  mutate(sims = invoke_map(f, params, n = 10))
```

Note that sims contains both double and integer vectors,
so it is not homogeneous. It is unlikely to cause problems,
since both types are numeric.

### 25.4.3 From multivalued summaries

summarise function only works with functions that return a single value.
So you can't use it with quantile:
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarise(q = quantile(mpg))
```

But you can wrap the result in a list !!
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarise(q = list(quantile(mpg)))
```

To use this with unnest, you should also capture the probabilities:
```{r}
probs <- c(0.01, 0.25, 0.5, 0.75, 0.99)

mtcars %>%
  group_by(cyl) %>%
  mutate(p = list(probs), q = list(quantile(mpg))) %>%
  unnest(c(p, q))
```

### 25.4.4 From a named list

Suppose you want to iterate over a column, and you also want to
iterate over the elements of a list which is in another column,
at the same time.
You can create a dataframe with one column being the singletons,
and the other column being the list using tibble::enframe.
```{r}
x <- list(
  a = 1:5,
  b = 3:4,
  c = 5:6
)

df <- enframe(x)
df
```

To iterate over the names and value in parallel:
```{r}
df %>%
  mutate(smry = map2_chr(name, value, ~stringr::str_c(.x, ": ", .y[1])))
```

### 25.4.5 Exercises

1. List all the functions that you can think of that take a atomic vector and return a list.

Many functions in the stringr package return a list:
str_split, str_match_all

map, map2 functions also return a list

2. Brainstorm useful summary functions that, like quantile(), return multiple values.

range function

3. What’s missing in the following data frame? How does quantile() return that missing piece? Why isn’t that helpful here?

We're missing probabilities of the quantiles. We can return these using
names = TRUE in the quantile function. But these names show up inside
each element of the q column. They're not available as a separate column.

quantile can return the names with names = TRUE, but we cannot have two
LHS values, only one. So, instead, we pull the names out after the
summarise with map(q, names). The problem with this is that all the
quantile names are in the names column per quantile value.

```{r}
name_quantile <- function(x) {
  names <- names(x)
  
  # This is ugly - but I don't yet know how to apply
  # the [[]] operator over a map. If x has a length
  # different from 5, this would be a problem.
  # The map would take care of it.
  values <- c(x[[1]], x[[2]], x[[3]], x[[4]], x[[5]])
  list(names, values)
}

mtcars %>%
  group_by(cyl) %>%
  summarise(
    q = list(quantile(mpg, names = TRUE))
  ) %>%
  mutate(separated = map(q, name_quantile)) %>%
  mutate(
    names = map(separated, function(x) { x[[1]] }),
    percentiles = map(separated, function(x) { x[[2]] })
  ) %>%
  unnest(c(names, percentiles))
```

4. What does this code do? Why might it be useful?

Usually, mtcars %>% group_by(cyl) %>% nest() will:
  
  * group by cyl
  * nest all columns related to each cyl value into a tibble,
    under the data column
    
mtcars %>% group_by(cyl) %>% summarise_all(list) does:

  * group by cyl
  * places a list of doubles into the cell for cyl row and mpg column.
    Does the same for each column (disp, hp, ...).
    All groups are converted to lists.
This is useful when applying transforms to grouped columns.
map can be applied to a single column for transforming it.
map2 can take two columns and apply a function that
  
  * uses values from both columns to create a new value
  * joins values from both columns into a new value
  
```{r}
x <- mtcars %>%
  group_by(cyl) %>%
  summarise_all(list)

x
```

## 25.5 Simplifying list-columns

Simplify the list-column into a regular column with atomic vector
as elements:

  * to get a single value (column), use mutate with:
    + map_lgl, map_int, map_dbl, map_chr
  * to get multiple values (columns), use unnest.
    This will repeat the rows multiple times
    
### 25.5.1 List to vector

```{r}
df <- tribble(
  ~x,
  letters[1:5],
  1:3,
  runif(5)
)

df %>%
  mutate(
    type = map_chr(x, typeof),
    length = map_int(x, length)
  )
```

This will allow you to filter out parts of a heterogeneous list-column.

You can use the map_* shortcuts - ex. map_chr(x, "apple") will
get you the string stored in the name apple for each element of x.
This helps pull apart nested lists into regular columns.
If an element is missing, and you don't want NULL returned,
use the .null argument.
```{r}
df <- tribble(
  ~x,
  list(a = 1, b = 2),
  list(a = 2, c = 4)
)

df %>%
  mutate(
    a = map_dbl(x, "a"),
    b = map_dbl(x, "b", .null = NA_real_)
  )
```

### 25.5.2 Unnesting

Unnesting works by repeating the elements of the regular columns
for each value of the list-column.
```{r}
tibble(x = 1:2, y = list(1:4, 1)) %>% unnest()
```

This means you cannot unnest two columns that contain
a different number of elements.
```{r}
df1 <- tribble(
  ~x,        ~y,     ~z,
  1, c("a","b"),    1:2,
  2,        "c",      3
)
df1
df1 %>% unnest(c(y, z)) # Works because 2 values in row 1 for 
                        # y and z. Also 1 value in row 2.
                        # So the x value can repeat 2 times for row 1
                        # and 1 times for row 2.
```

```{r}
df2 <- tribble(
  ~x,        ~y,     ~z,
  1,        "a",    1:2,
  2, c("b","c"),      3
)
df2
df2 %>% unnest(c(y, z)) # Gives a result by duplicating the second column.
                        # Is this unusual?
```

### 25.5.3 Exercises

1. Why might the lengths() function be useful for creating atomic vector columns from list-columns?

lengths function returns the length of each element of a list.
It can be useful shortcut to map_int(x, length).
Also, you can see if the lengths of the elements of a list
are equal.

2. List the most common types of vector found in a data frame. What makes lists different?

integer, double, character, logical, factor are usually found in
dataframes.
list is different because it contains more than one value,
and the type of each contained value can be different (heterogeneous).

## 25.6 Making tidy data with broom

The broom package has 3 different ways of turning models into tidy dataframes:

  1. broom::glance(model) returns a row for each model with 
     columns giving model quality (R^2, adjusted R^2, etc.)
  2. broom:tidy(model) returns a row for each coefficient of the model.
     Each column gives information about the estimate, or it's variability.
  3. broom::augment(model, data) returns a row for each data row,
     adding values like residuals and influence statistics.