---
title: "Ch_3_data_viz"
author: "Samir Gadkari"
date: "1/4/2021"
output: html_document
---

Hadley Wickham's [R for data science](https://r4ds.had.co.nz/data-visualisation.html), Chapter 3 problems and notes
```{r Load libraries}
library(ggplot2)
library(tidyverse)
library(maps)
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```
There are many points that are not seen in the plot above since they're
overlapping each other. You can use position = 'jitter' to shift those
points so that you see them all. This makes your plot less accurate
at small scales, but you can see point groupings much better at
the larger scales. As a shortcut, you can replace geom_point()
with geom_jitter() and just pass in the x and y aesthetics.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), position = 'jitter')
```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```

Not a very useful plot, since most of the class types
have all drive types:
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```

Using color aesthetic for class.
class is unordered, and so is color. This is fine.
We should not use an unordered aesthetic to map to an ordered feature,
and vice versa.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

This gives us a warning since we're using an ordered aesthetic (size)
to map to an unordered feature (class):
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Only 6 shapes can be used in a single plot.
Additional groups will be missing from the plot.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Manually setting color (aesthetic property of the geom).
All points are now in the color specified.
You can also set:
  shape = 10  or
  size = 3  (3mm)
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class), 
             color = 'blue', size = 3)
```

Mapping continuous variables to color, size.
Cannot map a continuous variable to shape - produces an error.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = year))
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = year))
```

Mapping the same variable to multiple aesthetics (shape, color, size) 
changes the shape, color, and size of the point groups.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, 
                           shape = class, color = class, size = class))
```

The stroke aesthetic modifies the width of the border
for shapes that have a border
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21, stroke = 1)
```

When you map an aesthetic to something other than a variable,
a True/False value is generated for each point based on the RHS
of the aesthetic, and the aesthetic is set based on the True/False
group it falls in.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

## Facets
Advantage of facets vs. color/shape/size is that you can see all values
without other data's values intruding on it.
The disadvantage is that it may be difficult to compare against
the other data points.
If there are lots of data points, facets would be preferred to clarify
the visualization. If not, color/shape/size may work better.
When using the facet_grid, put variable with more unique levels
in the columns, since laptop screens are more wide than tall.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2) # facet_wrap takes only 1 variable
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl) # facet_grid figures out how many rows/cols.
                        # First variable faceted along rows.
                        # Second variable faceted along columns.
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl) # To not facet along row or col, put a . there
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(cyl ~ .) # To not facet along row or col, put a . there
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ year) # When faceting on a continuous variable,
                       # only the minimum and maximum values
                       # of the variable are used for faceting.
```

Adding geom_smooth to fit the data with a smooth line.
The mapping was moved into the ggplot call since
it is common to both geom_point and geom_smooth.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()  # fit the data with a smooth line.
```

Use the linetype aesthetic to create separate geom_smooth plots.
You can use linetype to separate the geom_smooth plots,
and match those with the geom_point using color.
Notice that we've used linetype aesthetic on geom_smooth,
but not on geom_point (it wouldn't make sense anyway).
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(linetype = drv, color = drv)) +
  geom_point(mapping = aes(color = drv))
```

You can disable legend using show.legend = FALSE.
You have to do this for each geom.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(linetype = drv, color = drv), show.legend = FALSE) +
  geom_point(mapping = aes(color = drv), show.legend = FALSE)
```

## Specify different data for each layer
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == 'subcompact'), # select only 
                                                         # subcompact cars.
              se = FALSE) # SE shows the 95% confidence interval, and is the 
                          # gray shaded area around the smooth fit line.
```

```{r}
ggplot(data = filter(mpg, drv == 'f'), mapping = aes(x = displ, y = hwy)) + 
  geom_point()
```
```{r}
ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) + 
  geom_boxplot()
```
```{r}
ggplot(data = mpg, mapping = aes(x = hwy, y = displ)) +
  geom_area()
```

We're grouping by color on drv, thus geom_smooth creates multiple
fits
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

## Various ways of generating plots
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(data = filter(mpg, drv == '4'), se = FALSE) +
  geom_smooth(data = filter(mpg, drv == 'f'), se = FALSE) +
  geom_smooth(data = filter(mpg, drv == 'r'), se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = drv)) + 
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point()
```

## Statistical Transformations on the diamonds data
Here we calculate the count of the number in each category of cut.
Bar charts, histograms, frequency polygons bin the data and plot bin counts.
Boxplots calculate percentiles and plot them.
Smoothers calculate a best-fit smooth line and plot it.

The algorithm to calculate a new value for a plot is the stat.
You can find the stat by looking at help on geom_bar etc.
Since there is a 1-to-1 correspondence between the geom and the stat,
you can even use the stat function to do the plotting.
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```
```{r}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

You can change the stat function used, by specifying y in the aesthetics.
Below, we're using the identity stat to get the prices per cut.
```{r}
demo <- tribble(
  ~cut,       ~freq,
  'fair',     101,
  'good',     202,
  'v good',   230,
  'premium',  480,
  'ideal',    820
)
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), 
           stat = "identity") # Change default stat associated with this geom.
                              # Change default geom associated with stat
                              # when using stat_* functions using geom=...
```

To show a bar chart of proportion, use the proportion stat as an aesthetic.
You can map an aesthetic to:

  * a data variable (ex. x = cut)
  * after data is transformed by the stat
    (ex. y = after_stat(count / max(count))).
    Here, the proportion is calculated by the stat.
  * after data has been transformed and mapped by the plot scales
    (ex. after_scale = alpha(fill, 0.4))
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, 
                         y = stat(prop), # plot proportions. Not a stat?
                         group = 1))     # group = 1 says add all proportions.
                                         # Without this, geom_bar does not know
                                         # what to do with the individual
                                         # proportions.
```

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop),
                         group = color)) # ??? is this correct?
# Without the group variable, geom_bar does not know what to do with all
# the proportions. group = color tells it to split all colors for each cut.
# Is this right ???
```

## Position adjustments
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
To plot each bar to be equal height, use position = 'fill'.
This will allow you to compare the values of each clarity proportion
across the cut.
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = 'fill')
```
You can also place each clarity bar as a separate bar for each cut using
fill = 'dodge'
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = 'dodge')
```

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_point()
```
This plot is not showing all points, let's use position = 'jitter'.
The amount of jitter is controlled by the width/height parameter of geom_jitter.
The default width/height is 0.4 of the data resolution.
The jitter is added in both positive and negative directions, so the total
jitter is by default 0.8.
Categorical data is aligned on integers, so a value of 0.5 will
move points into the other levels.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = 'jitter')
```
You can now see how there are clumps at:

  * cty/hwy around 7
  * cty around 12, hwy around 10
  * cty around 13, hwy around 15
  * cty around 17, hwy around 26
  * cty around 22, hwy around 29

An alternative to geom_jitter is geom_count which counts the number of points
at each location and draws a shape of a proportional size:
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```

The default position adjustment for the geom_boxplot is 'dodge2'.
With position changed to 'dodge', or 'identity',
the plot does not change.
With position changed to 'fill', we get an error.
```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = displ), position='dodge')
```

To show a summary of the min, max, median values per category:
```{r}
ggplot(data = diamonds) +
  stat_summary(  # stat_summary uses geom pointrange
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

Comparing geom_bar and geom_col:

  * geom_bar makes height of bar = count of cases in each group
    (using stat_count)
  * geom_col makes height of bar = value of data
    (using stat_identity)

## Coordinate systems

Usually, x axis is horizontal, and y axis is vertical.
To flip them, use coord_flip().
This is useful when creating plots that have a large number of columns,
and cannot fit in the display width. Also helps when the labels are longer.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
```

To set the aspect ratio correctly for maps, use coord_quickmap()
```{r}
nz <- map_data('nz')
ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = 'white', color = 'black')
ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = 'white', color = 'black') +
  coord_quickmap()
```

coord_polar() to use polar coordinates.
```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(
    mapping = aes(x = cut, fill = cut),
    show.legend = FALSE,
    width = 1
  ) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL) # labs() can modify axis, legend and plot labels.
                           # x = '', y = '' modifies the axis labels
                           # title/subtitle = '' explains main findings
                           # caption = '' gives information on data source
                           # tag = '' identifies a plot to differentiate it
                           #          among multiple plots.
                           # All of these can be set to waiver(). This indicates
                           # that the default value should be used instead.

bar + coord_flip()
bar + coord_polar()
```

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline()
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed() # coord_fixed() sets the ratio of the  screen distance
                # of 1 unit on the y axis to 1 unit on the x axis.
                # The default ratio in the coord_fixed() function is 1.
                # This allows us to compare two different plots that we create
                # with the same ratio.
                # Without this fix, the ratios could be different among
                # a set of graphs.
```

