---
title: "Ch_3_data_viz"
author: "Samir Gadkari"
date: "1/4/2021"
output: html_document
---

Hadley Wickham's [R for data science](https://r4ds.had.co.nz/data-visualisation.html), Chapter 3 problems and notes
```{r Load libraries}
library(ggplot2)
library(tidyverse)
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```

Not a very useful plot, since most of the class types
have all drive types:
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```

Using color aesthetic for class.
class is unordered, and so is color. This is fine.
We should not use an unordered aesthetic to map to an ordered feature,
and vice versa.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

This gives us a warning since we're using an ordered aesthetic (size)
to map to an unordered feature (class):
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Only 6 shapes can be used in a single plot.
Additional groups will be missing from the plot.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Manually setting color (aesthetic property of the geom).
All points are now in the color specified.
You can also set:
  shape = 10  or
  size = 3  (3mm)
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class), 
             color = 'blue', size = 3)
```

Mapping continuous variables to color, size.
Cannot map a continuous variable to shape - produces an error.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = year))
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = year))
```

Mapping the same variable to multiple aesthetics (shape, color, size) 
changes the shape, color, and size of the point groups.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, 
                           shape = class, color = class, size = class))
```

The stroke aesthetic modifies the width of the border
for shapes that have a border
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21, stroke = 1)
```

When you map an aesthetic to something other than a variable,
a True/False value is generated for each point based on the RHS
of the aesthetic, and the aesthetic is set based on the True/False
group it falls in.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

## Facets
Advantage of facets vs. color/shape/size is that you can see all values
without other data's values intruding on it.
The disadvantage is that it may be difficult to compare against
the other data points.
If there are lots of data points, facets would be preferred to clarify
the visualization. If not, color/shape/size may work better.
When using the facet_grid, put variable with more unique levels
in the columns, since laptop screens are more wide than tall.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2) # facet_wrap takes only 1 variable
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl) # facet_grid figures out how many rows/cols.
                        # First variable faceted along rows.
                        # Second variable faceted along columns.
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl) # To not facet along row or col, put a . there
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(cyl ~ .) # To not facet along row or col, put a . there
```
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ year) # When faceting on a continuous variable,
                       # only the minimum and maximum values
                       # of the variable are used for faceting.
```

Adding geom_smooth to fit the data with a smooth line.
The mapping was moved into the ggplot call since
it is common to both geom_point and geom_smooth.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()  # fit the data with a smooth line.
```

Use the linetype aesthetic to create separate geom_smooth plots.
You can use linetype to separate the geom_smooth plots,
and match those with the geom_point using color.
Notice that we've used linetype aesthetic on geom_smooth,
but not on geom_point (it wouldn't make sense anyway).
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(linetype = drv, color = drv)) +
  geom_point(mapping = aes(color = drv))
```

You can disable legend using show.legend = FALSE.
You have to do this for each geom.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(linetype = drv, color = drv), show.legend = FALSE) +
  geom_point(mapping = aes(color = drv), show.legend = FALSE)
```

## Specify different data for each layer
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == 'subcompact'), # select only 
                                                         # subcompact cars.
              se = FALSE) # SE shows the 95% confidence interval, and is the 
                          # gray shaded area around the smooth fit line.
```

```{r}
ggplot(data = filter(mpg, drv == 'f'), mapping = aes(x = displ, y = hwy)) + 
  geom_point()
```
```{r}
ggplot(data = mpg, mapping = aes(x = drv, y = hwy)) + 
  geom_boxplot()
```
```{r}
ggplot(data = mpg, mapping = aes(x = hwy, y = displ)) +
  geom_area()
```

We're grouping by color on drv, thus geom_smooth creates multiple
fits
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

## Various ways of generating plots
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(data = filter(mpg, drv == '4'), se = FALSE) +
  geom_smooth(data = filter(mpg, drv == 'f'), se = FALSE) +
  geom_smooth(data = filter(mpg, drv == 'r'), se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = drv)) + 
  geom_smooth(se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point()
```

## Statistical Transformations