---
title: "Ch_5_data_transforms"
author: "Samir Gadkari"
date: "1/5/2021"
output: html_document
---

Hadley Wickham's [R for data science](https://r4ds.had.co.nz/data-visualisation.html), Chapter 5 problems and notes
```{r Load libraries}
library(nycflights13)
library(tidyverse)
```

Since this is a large dataset, only a few rows are printed when you type
flights in the console. To see all the rows, type View(flights) in the console.
This will open the RStudio viewer.

## dplyr basics

Main functions in dplyr that allow you to manipulate the data:

  * filter: pick observations by their name
  * arrange: reorder 
  * select: pick variables by their names
  * mutate: create new variables with functions of existing variables
  * summarise: collapse many values into a summary
These functions used in conjunction with group_by (which changes the function
scope to group level) allow all kinds of data manipulation.

For all these functions:

1. First argument is a dataframe;
2. subsequent arguments describe what to do with the dataframe
3. resulting in a new dataframe

dplyr functions never modify the dataframe.
You have to set it to a variable to capture the updated values.

## Filter rows with filter()
To print out the output of any code, surround it with parentheses.
The line below saves the filter output to jan1 and prints out the
value in jan1.
```{r}
(jan1 <- filter(flights, month == 1, day == 1)) # Multiple arguments are combined 
                                                # with the and operation.
                                                # Use logical operands &, |, !
                                                # to combine using and, or, not.
```
When filtering, if a row produces FALSE, or NA, the row is dropped.
To preserve NA values, you have to ask for them explicitly.
```{r}
(df <- tibble(x = c(1, NA, 3)))
filter(df, x > 1)
(filter(df, is.na(x) | x > 1))
```

### Comparisons

When comparing two floating point numbers, use near()
```{r}
near(sqrt(2) ^ 2, 2)
```

### Logical operations

```{r}
filter(flights, month == 11 | month == 1) # Either month 11 or 1
```

The same can be achieved with the %in% operator
```{r}
filter(flights, month %in% c(11, 1))
```
You can compare two tibbles. The value of each tibble will be compared
individually:
```{r}
x <- filter(flights, month == 11 | month == 1)
x
```

```{r}
x <- filter(flights, month == 11 | month == 1)
y <- filter(flights, month %in% c(11, 1))
all_equal(x, y)
(x == y)[1, ]
```

### Missing values
Any calculation that has a value of NA will produce an NA result.
NA stands for "not available". Even NA == NA will result in NA.
To determine if a value is NA, use is.na().
Note that is.na will find NaN as well as NA.
is.nan will find only NaN values.
```{r}
x <- NA
# y <- 32
is.na(x)
# is.na(y)
```
Some calculations with NA in them will still work.
This is because of shortcuts that the computer uses to do the calculation.
```{r}
(NA ^ 0)     # Anything raised to 0 is 1
(NA | TRUE)  # Anything or-ed with TRUE is TRUE
(FALSE & NA) # Anything an-ded with FALSE is FALSE
(NA * 0)     # Output is NA, because 0 can be a number close to but not 0.
             # So any tiny tiny number times NA should give NA.
```

### 5.2.4 Exercises
Filtering exercises:
```{r}
(filter(flights, sched_arr_time - arr_time >= 120))
(filter(flights, dest == 'HOU' | dest == 'IAH'))
(filter(flights, carrier %in% c('UA', 'AA', 'DL')))

(filter(flights, month %in% 7:9))       # One way to do this, and
(filter(flights, between(month, 9, 9))) # another way to do it.

(filter(flights, (sched_arr_time - arr_time >= 120) & dep_delay < 0))
(filter(flights, (dep_delay > 60) & (sched_arr_time - arr_time < 30)))
(filter(flights, dep_time <= 600))
```

```{r}
# How many flights have a missing dep_time?
(filter(flights, is.na(dep_time))) # There are 8255 rows with NA in dep_time.
(sum(is.na(flights$dep_time)))     # Another way to do this without printing
                                   # the tibble and reading number of rows.

# Find number of missing values per variable
(colSums(is.na(flights)))          # Best way to see number of missing values
                                   # per variable.
```

