---
title: "Ch_7_exploratory_data_analysis.Rmd"
author: "Samir Gadkari"
date: "1/12/2021"
output: html_document
---

Hadley Wickham's [R for data science](https://r4ds.had.co.nz/exploratory-data-analysis.html), Chapter 7 problems and notes
```{r Load libraries}
library(tidyverse)
```

Since this is a large dataset, only a few rows are printed when you type
flights in the console. To see all the rows, type View(flights) in the console.
We're going to explore our data through:
  * variations within the variables
  * covariations between variables

## Variation

You can use bar plots to find variations in categorical variables:
```{r}
# Variation as a plot:
diamonds %>%
  ggplot(aes(x = cut)) +
  geom_bar()

# Variation as a table:
diamonds %>%
  count(cut) %>%           # counts the number of each type
  mutate(prop = n/sum(n))  # gives the proportion of each type
```

You can use histograms to find variations in continuous data:
```{r}
diamonds %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.5) # You should explore data with different binwidths

diamonds %>%
  count(cut_width(carat, 
                  0.5)) %>%  # width is required for cut_width
  mutate(prop = n / sum(n))
```

To overlay multiple histograms, use geom_freqpoly(). It uses a different
colored line for each histogram, and so is easier to understand.
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, color = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

### Typical values

  * Which values are the most common? Why?
  * Which values are rare? Why? Does that match your expectations?
  * Can you see any unusual patterns? What might explain them?
